---


---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gplots)
library(tidyverse)
```
Import Data
```{r}
#wd = "D:/ Data Analysis/Bike sharing data"
#setwd(wd)

bikeHour = read.csv("hour.csv")
bikeDay = read.csv("day.csv")
```
Check the data
```{r}
str(bikeHour)

```
Data Cleaning
```{r}
##############################################################################
##############################  FILTERING ####################################
##############################################################################



############################# Hourly Dataset #################################

## Missing Values
missingHour = sum(is.na(bikeHour)) 
missingDay  = sum(is.na(bikeDay))
# there are no missing values

## Outliers - CASUAL----------------------------------------------------------
Array = bikeHour$casual
boxplot(Array) # Boxplot

# Probability Distribution Function
A = mean(Array)  # mean
B = sd(Array)    # standard deviation
y = dnorm(Array, A, B)         # PDF

plot(Array,y, main = "PDF")  # PDF of the Array

# Inter Quartile Range
summary(Array)
iqr = IQR(Array) # interquartile range
lb  = quantile(Array, 0.25) - 1.5*iqr  # lower bound
ub  = quantile(Array, 0.75) + 1.5*iqr  # upper bound

iqr_bikeHourCasual = bikeHour[Array >= lb & Array <= ub,] # data only within IQR


## Zero Values
# could correlate with other conditions ie. weather

zeroes_bikeHourCasual = bikeHour[Array == 0,]
omitZeroes_bikeHourCasual = bikeHour[Array != 0,]

# IQR of non-zero data
nonzero = Array[Array != 0]
iqr_omitZeroes = IQR(nonzero) 
lb  = quantile(nonzero, 0.25) - 1.5*iqr_omitZeroes  # lower bound
ub  = quantile(nonzero, 0.75) + 1.5*iqr_omitZeroes  # upper bound

iqr_omitZeroes_bikeHourCasual = omitZeroes_bikeHourCasual[nonzero >= lb & nonzero <= ub,]

## Outliers - Registered ------------------------------------------------------
Array = bikeHour$registered
boxplot(Array) # Boxplot

# Probability Distribution Function
A = mean(Array)  # mean
B = sd(Array)    # standard deviation
y = dnorm(Array, A, B)         # PDF

plot(Array,y, main = "PDF")  # PDF of the Array

# Inter Quartile Range
summary(Array)
iqr = IQR(Array) # interquartile range
lb  = quantile(Array, 0.25) - 1.5*iqr  # lower bound
ub  = quantile(Array, 0.75) + 1.5*iqr  # upper bound

iqr_bikeHourRegistered = bikeHour[Array >= lb & Array <= ub,] # data only within IQR


## Zero Values
# could correlate with other conditions ie. weather

zeroes_bikeHourRegistered = bikeHour[Array == 0,]
omitZeroes_bikeHourRegistered = bikeHour[Array != 0,]

# IQR of non-zero data
nonzero = Array[Array != 0]
iqr_omitZeroes = IQR(nonzero) 
lb  = quantile(nonzero, 0.25) - 1.5*iqr_omitZeroes  # lower bound
ub  = quantile(nonzero, 0.75) + 1.5*iqr_omitZeroes  # upper bound

iqr_omitZeroes_bikeHourRegistered = omitZeroes_bikeHourRegistered[nonzero >= lb & nonzero <= ub,]

## Outliers - General----------------------------------------------------------

omitZeroes_bikeHour = bikeHour[bikeHour$casual != 0,]
omitZeroes_bikeHour = omitZeroes_bikeHour[omitZeroes_bikeHour$registered != 0,]


############################# Daily Dataset #################################


## Outliers - CASUAL----------------------------------------------------------
Array = bikeDay$casual
boxplot(Array) # Boxplot

# Probability Distribution Function
A = mean(Array)  # mean
B = sd(Array)    # standard deviation
y = dnorm(Array, A, B)         # PDF

plot(Array,y, main = "PDF")  # PDF of the Array

# Inter Quartile Range
summary(Array)
iqr = IQR(Array) # interquartile range
lb  = quantile(Array, 0.25) - 1.5*iqr  # lower bound
ub  = quantile(Array, 0.75) + 1.5*iqr  # upper bound

iqr_bikeDayCasual = bikeDay[Array >= lb & Array <= ub,] # data only within IQR


## Zero Values
# could correlate with other conditions ie. weather

zeroes_bikeDayCasual = bikeDay[Array == 0,]
omitZeroes_bikeDayCasual = bikeDay[Array != 0,]

# IQR of non-zero data
nonzero = Array[Array != 0]
iqr_omitZeroes = IQR(nonzero) 
lb  = quantile(nonzero, 0.25) - 1.5*iqr_omitZeroes  # lower bound
ub  = quantile(nonzero, 0.75) + 1.5*iqr_omitZeroes  # upper bound

iqr_omitZeroes_bikeDayCasual = omitZeroes_bikeDayCasual[nonzero >= lb & nonzero <= ub,]

## Outliers - Registered ------------------------------------------------------
Array = bikeDay$registered
boxplot(Array) # Boxplot

# Probability Distribution Function
A = mean(Array)  # mean
B = sd(Array)    # standard deviation
y = dnorm(Array, A, B)         # PDF

plot(Array,y, main = "PDF")  # PDF of the Array

# Inter Quartile Range
summary(Array)
iqr = IQR(Array) # interquartile range
lb  = quantile(Array, 0.25) - 1.5*iqr  # lower bound
ub  = quantile(Array, 0.75) + 1.5*iqr  # upper bound

iqr_bikeDayRegistered = bikeDay[Array >= lb & Array <= ub,] # data only within IQR


## Zero Values
# could correlate with other conditions ie. weather

zeroes_bikeDayRegistered = bikeDay[Array == 0,]
omitZeroes_bikeDayRegistered = bikeDay[Array != 0,]

# IQR of non-zero data
nonzero = Array[Array != 0]
iqr_omitZeroes = IQR(nonzero) 
lb  = quantile(nonzero, 0.25) - 1.5*iqr_omitZeroes  # lower bound
ub  = quantile(nonzero, 0.75) + 1.5*iqr_omitZeroes  # upper bound

iqr_omitZeroes_bikeDayRegistered = omitZeroes_bikeDayRegistered[nonzero >= lb & nonzero <= ub,]

## Outliers - General----------------------------------------------------------

omitZeroes_bikeDay = bikeDay[bikeDay$casual != 0,]
omitZeroes_bikeDay = omitZeroes_bikeDay[omitZeroes_bikeDay$registered != 0,]
```

Get rid of "instant"
```{r}
bikeHour <- bikeHour[,-1]
bikeDay <- bikeDay[,-1]
```
Transform variables into factor and rename levels for
```{r}
bikeHour <- bikeHour %>%
  mutate(across(c(season,yr,mnth,workingday,holiday,weekday,weathersit),as.factor))

bikeHour$season <- recode_factor(bikeHour$season, "1" = "Spring", "2"= "Summer", "3" = "Fall", "4" = "Winter") 
bikeHour$workingday <- recode_factor(bikeHour$workingday, "1" = "Workday", "0" = "Weekend/Holiday")
bikeHour$weathersit <- recode_factor(bikeHour$weathersit, "1" = "Clear", "2"= "Mist/Cloudy", "3" = "Light Rain/Snow", "4" = "Heavy Rain/Storm")
bikeHour$holiday <- recode_factor(bikeHour$holiday, "1" = "Holiday", "0" = "Not Holiday")
bikeHour$mnth <- recode_factor(bikeHour$mnth, "1" = "January", "2" = "February", "3" = "March", "4" = "April", "5" = "May", "6" = "June", "7" = "July", "8" = "August", "9" = "September", "10" = "October", "11" = "November", "12" = "December")
str(bikeHour)


```

Ration of casual and registered
```{r}
bikeHour <- mutate(bikeHour, Ratio = (registered+1)/(casual+1))
```

Make a new factor variable morning/afternon/evening/night
mor (1) = 5:00 am to 11:00 am, aft(2) = 11:00 am to 5:00pm, evn(3) = 5:00 pm to 11:00 pm, nig(4) = 11:00pm to 5:00 am
```{r}
bikeHour <- mutate(bikeHour, TimeOfDay = case_when(
    hr %in% c(5,11) ~ "Morning",
    hr %in% c(12,17) ~ "Afternoon",
    hr %in% c(18,22) ~ "Evening",
    TRUE ~ "Night"))
bikeHour$TimeOfDay <- as.factor(bikeHour$TimeOfDay)
bikeHour$hr<- as.factor(bikeHour$hr)

```
view(bikeHour)
```{r}
plot_trend_day_hour <- ggplot(bikeHour, aes(x = weekday, y = cnt, fill = hr)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("red", "blue", "green", "yellow", "violet", "pink", "purple", "black",
                               "grey", "brown", "cyan", "orange", "beige", "light pink", "gold",
                               "sienna1", "hotpink2", "aquamarine", "wheat", "plum", "khaki", "azure",
                               "sky blue", "tan")) +
  labs(title = "Bike Rent Trend Every Hour",
       subtitle = "Group by Day",
       x = "Day Name",
       y = "Freq",
       fill = "") +
  theme_minimal() +
  theme(legend.position = "bottom",
        legend.title = element_blank())

plot_trend_day_hour
```
```{r}
library(ggplot2)

# Create a grouped or stacked bar chart to compare "casual" and "registered" bike usage
ggplot(bikeHour, aes(x = mnth)) +
  geom_bar(aes(y = casual, fill = "blue"), stat = "identity", position = "dodge", width = 0.4) +
  geom_bar(aes(y = registered, fill = "green"), stat = "identity", position = "dodge", width = 0.4) +
  scale_fill_manual(values = c("Casual" = "blue", "Registered" = "green")) +
  labs(
    title = "Monthly Bike Usage Comparison",
    x = "Month",
    y = "Number of Bikes",
    fill = "Category"
  ) +
  theme_minimal() +
  theme(legend.position = "top")
```
```{r}
#doubt
library(ggplot2)
 
ggplot(bikeHour, aes(x = mnth)) +
  geom_line(aes(y = casual), color = "blue", linetype = "solid") +
  geom_line(aes(y = registered), color = "red", linetype = "dashed") +
  labs(
    title = "Plot with One X-Variable and Two Y-Variables",
    x = "X-Axis Label",
    y = "Y-Axis Label"
  ) +
  theme_minimal()
```

```{r}
boxplot(temp ~ season,
        data = bikeHour,
        xlab = "season",
        ylab = "temp",
        main = "Temperature by Season",
        col = "blue")
```
# create a beanplot for number of bike rents per each weather condition
#install.packages("beanplot")
library("beanplot")
require("beanplot")
require("RColorBrewer")
beanplot <- lapply(brewer.pal(6, "Set3"),
                    function(x){return(c(x, "black", "gray", "red"))})
beanplot(cnt ~ weathersit,
         data = bikeHour,
         main = "Bike Rents by Weather Condition",
         xlab = "Weather Condition",
         ylab = "Number of rentals",
         col = beanplot,
         lwd = 1,
         what = c (1,1,1,0),
         log = ""
         )
```{r}
#weather vs bike rentals 
library(ggplot2)

# Set the colors for the bars
colors <- brewer.pal(6, "Set3")

# Create a bar chart
ggplot(bikeHour, aes(x = weathersit, y = cnt, fill = weathersit)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = colors) +
  labs(
    title = "Bike Rents by Weather Condition",
    x = "Weather Condition",
    y = "Number of Rentals"
  ) +
  theme_minimal()
```
  
#casual vs registered
```{r}
#install.packages("plotly")
library(plotly)
seasonality <- plot_ly()

seasonality <- add_trace(seasonality, x = ~bikeHour$season, y = ~bikeHour$casual, name = 'Casual', 
                 type = 'scatter', mode = 'lines', line = list(color = 'black', width = 4))
seasonality<- add_trace(seasonality, x = bikeHour$season, y = bikeHour$registered, name = 'Registered', 
                 type = 'scatter', mode = 'lines', line = list(color = 'red', width = 4))

seasonality <- layout(seasonality, title = 'Casual vs. Registered rentals per season', xaxis = list(title = 'Season'), 
              yaxis = list(title = 'Rentals'))

seasonality
```

```{r}
# Load the dplyr library
library(dplyr)

# Convert "season" to a factor if it's not already
bikeHour$season <- as.factor(bikeHour$season)

# Group by 'season' and sum 'registered' and 'casual' columns
season_registered_data <- bikeHour %>%
  group_by(season) %>%
  summarize(registered = sum(registered))

season_casual_data <- bikeHour %>%
  group_by(season) %>%
  summarize(casual = sum(casual))

# Normalize 'registered' and 'casual' values
season_registered_data <- season_registered_data %>%
  mutate(registered = registered / sum(registered))

season_casual_data <- season_casual_data %>%
  mutate(casual = casual / sum(casual))

# Reset the index
season_registered_data <- season_registered_data %>%
  ungroup()

season_casual_data <- season_casual_data %>%
  ungroup()
```

```{r}
# Load the plotly library
library(plotly)

# Assuming you have 'season_casual_data' and 'season_registered_data' data frames
# Create the plot
fig <- plot_ly()

fig <- add_trace(fig, data = season_casual_data, x = ~season, y = ~casual, name = 'Casual', type = 'scatter', mode = 'lines', line = list(color = 'red', width = 4))
fig <- add_trace(fig, data = season_registered_data, x = ~season, y = ~registered, name = 'Registered', type = 'scatter', mode = 'lines', line = list(color = 'black', width = 4))

fig <- layout(fig, title = 'Casual vs. Registered rentals per season', xaxis = list(title = 'Season'), yaxis = list(title = 'Rentals'))

# Show the plot
fig
```


```{r}
library(ggplot2)

# Assuming df_casual_avg is your data frame
# Create a grouped bar plot
plot_trend_casual <- ggplot(bikeHour, aes(x = season, y = casual, fill = factor(yr))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Casual rental averages per season and year",
    x = "Season",
    y = "Casual Average",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("#88D1CC", "#F6DCAD")) +  # Define colors for the years
  theme_minimal()

# Show the plot
print(plot_trend_casual)
```

```{r}
# Create a grouped bar plot
plot_trend_registered <- ggplot(bikeHour, aes(x = season, y = registered, fill = factor(yr))) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "registered rental averages per season and year",
    x = "Season",
    y = "registered Average",
    fill = "Year"
  ) +
  scale_fill_manual(values = c("#88D1CC", "#F6DCAD")) +  # Define colors for the years
  theme_minimal()

# Show the plot
print(plot_trend_registered)
```


```{r}
t1 <- cor.test(bikeHour$atemp[bikeHour$yr == 1],
               bikeHour$cnt[bikeHour$yr == 1])
t1
t2 <- cor.test(bikeHour$atemp[bikeHour$yr == 0],
               bikeHour$cnt[bikeHour$yr == 0])
t2
```
library(ggplot2)

# Create a scatter plot with regression lines
scatter_plot <- ggplot(bikeHour, aes(x = atemp, y = cnt, color = factor(yr))) +
  geom_point(size = 3) +
  geom_smooth(method = 'lm', se = FALSE, size = 1, linetype = 'solid') +  # Add regression lines
  labs(
    title = "Temperature vs. Count with Regression Lines",
    x = "Temperature",
    y = "Number of Rents",
    color = "Year"
  ) +
  scale_color_manual(values = c("0" = "red", "1" = "black")) +  # Define colors for the years
  theme_minimal()

# Show the scatter plot with regression lines
print(scatter_plot)

```{r}
#install.packages("gridExtra")
library(gridExtra)
# Load the required library
library(ggplot2)

# Create a vector for histogram titles
vec <- c("Clear Weather", "Cloudy Weather", "Rainy Weather", "Thunderstorm Weather")

# Create a list to store the plots
plot_list <- list()

# Create histograms with median and mean lines for each weather condition
for (i in 1:4) {
  name_i <- vec[i]
  data_subset <- subset(bikeHour, weathersit == i)
  
  # Create a histogram
  histogram <- ggplot(data_subset, aes(x = cnt)) +
    geom_histogram(binwidth = 20, fill = "yellow3", color = "black") +
    labs(
      title = name_i,
      x = "Number of Rents",
      y = "Frequency"
    ) +
    geom_vline(aes(xintercept = median(cnt)),
               color = "black",
               linetype = "dashed",
               size = 1) +
    geom_vline(aes(xintercept = mean(cnt)),
               color = "blue",
               linetype = "dashed",
               size = 1)
  
  plot_list[[i]] <- histogram
}

# Combine the histograms in a 2x2 grid
grid.arrange(grobs = plot_list, ncol = 2)

# Reset the layout
dev.off()
histogram
```
```{r}
beanplot(cnt ~ holiday,
         data = bikeHour,
         main = "Bike Rents by Type of a Day",
         xlab = "Type of Day",
         ylab = "Number of rents",
         col = beanplot,
         lwd = 1,
         what = c(1,1,1,0),
         log = ""
         )
```
```{r}
library(ggplot2)

# Create a bar graph
bar_graph <- ggplot(bikeHour, aes(x = factor(holiday))) +
  geom_bar(fill = "blue") +
  labs(
    title = "Bike Rents by Type of a Day",
    x = "Type of Day",
    y = "Number of rents"
  ) +
  theme_minimal()

# Show the bar graph
print(bar_graph)
```

